Eco-evolutionary dynamics of finite populations from first principles

### 2023-02-21 ind_files/99_Appendix_C.tex
Simplify and correct derivation of variance Price eqn

&+ \frac{1}{KN^2_{K}(t)}\left(\sum\limits_{i=1}^{m}f^2_iA_i^+ - \overline{f}^2\sum\limits_{i=1}^{m}A_i^+\right)dt\\

@@

\frac{1}{KN^2_{K}(t)}\left(\sum\limits_{i=1}^{m}f^2_iA_i^+ - \overline{f}^2\sum\limits_{i=1}^{m}A_i^+\right)dt &= \frac{1}{KN_{K}}\sum\limits_{i=1}^{m}f_i^2\left(\tau_ip_i + \mu Q_i(\mathbf{p})\right) - \frac{\overline{f}^2}{KN_{K}}\sum\limits_{i=1}^{m}\left(\tau_ip_i + \mu Q_i(\mathbf{p})\right)\\
&= \frac{1}{KN_K}\sum\limits_{i=1}^{m}\left(f_i^2 - \overline{f}^2\right)\left(\tau_ip_i + \mu Q_i(\mathbf{p})\right)\\
&= \frac{1}{KN_K}\left(\textrm{Cov}(\tau,f^2)+\mu \sum\limits_{i=1}^{m}\left(f_i^2 - \overline{f}^2\right)Q_i(\mathbf{p})\right)

@@

Substituting this into \eqref{intermediate_1_for_variances} and using $M_{\sigma^2_f}(\mathbf{p},N_K)$ to denote the contributions of all the mutational terms (\emph{i.e.} all terms with a $\mu$ factor) for notational brevity, we obtain

@@

d\sigma^2_{f} &= \textrm{Cov}(w,f^2 - 2\overline{f}f)dt - \frac{1}{KN_K}\left(\textrm{Cov}(\tau,2f^2 - 2\overline{f}f)\right)dt\\
&+ 2\textrm{Cov}\left(\frac{\partial f}{\partial t},f\right)dt + M_{\sigma^2_f}(\mathbf{p},N_K)dt + dB_{\sigma^2_{f}}

@@

&- \frac{1}{KN_K}\left[ \ \textrm{Cov}\left(\tau,(f - \overline{f})^2\right) + \textrm{Cov}\left(\tau,f^2 - {\left(\overline{f}\right)}^2\right) \ \right]dt\\[12pt]

@@

&= {\left(\overline{f}\right)}^2\overline{w} - \overline{w}{\left(\overline{f}\right)}^2 = 0\label{cov_term_1_for_variances}
and
\textrm{Cov}\left(\tau,f^2 - {\left(\overline{f}\right)}^2\right) &= \overline{\tau\left(f^2 - {\left(\overline{f}\right)}^2\right)} - \overline{\tau}\left(\overline{f^2 - {\left(\overline{f}\right)}^2}\right)\\
&= \overline{\tau f^2} - \overline{\tau}{\left(\overline{f}\right)}^2  - \overline{\tau}\overline{f^2} + \overline{\tau}{\left(\overline{f}\right)}^2\\
&=\overline{\tau f^2} - \overline{\tau}\overline{f^2} = \textrm{Cov}(\tau,f^2)\label{cov_term_2_for_variances}
Substituting \eqref{cov_term_1_for_variances} and \eqref{cov_term_2_for_variances} into \eqref{intermediate_2_for_variances}, we thus arrive at:

@@

\label{intermediate_3_for_variances}

@@

d\sigma^2_{f} &= \textrm{Cov}\left(w,(f - \overline{f})^2\right)dt\\
&- \frac{1}{KN_K}\left[ \ \textrm{Cov}\left(\tau,(f - \overline{f})^2\right) + \textrm{Cov}(\tau,f^2) \ \right]dt\\

@@

Covariance with $(f-\overline{f})^2$ is a measure of covariance with the existing `spread' of $f_i$ around the population mean value $\overline{f}$ and is thus readily understood to represent existing variation. However, the $\textrm{Cov}(\tau,f^2)$ term is difficult to interpret since it depends on the square of $f$, which may not have any direct biological interpretation. We will convert this into a form whose interpretation is more clear by repeatedly adding and subtracting a quantity from our expression (these quantities are shown in red below for clarity), noting that this leaves the net value unchanged. We have:
\textrm{Cov}(\tau, f^2) &= \overline{\tau f^2} - \overline{\tau}\overline{f^2}\nonumber\\
&= \overline{\tau f^2} {\color{red} + \overline{\tau{\left(\overline{f}\right)}^2} - \overline{\tau f \overline{f}}}  - \overline{\tau}\overline{f^2} \nonumber\\
&= \overline{\tau f^2} + \overline{\tau{\left(\overline{f}\right)}^2} - 2\overline{\tau f \overline{f}}\nonumber\\
&= \overline{\tau\left(f-\overline{f}\right)^2}\nonumber\\
&= \overline{\tau\left(f-\overline{f}\right)^2} {\color{red}- \overline{\tau}{\left(\overline{f}\right)}^2 + 2 \overline{\tau}\overline{f\overline{f}} - \overline{\tau}{\left(\overline{f}\right)}^2}\nonumber\\
&= \overline{\tau\left(f-\overline{f}\right)^2} {\color{red} - \overline{\tau}\overline{f^2}} -\overline{\tau}\overline{{\left(\overline{f}\right)}^2} + 2 \overline{\tau}\overline{f\overline{f}} {\color{red} + \overline{\tau}\overline{f^2}} - \overline{\tau}{\left(\overline{f}\right)}^2\nonumber\\
&=  \overline{\tau\left(f-\overline{f}\right)^2} - \overline{\tau}\overline{\left(f-\overline{f}\right)^2} + \overline{\tau}\left(\overline{f^2} - {\left(\overline{f}\right)}^2\right)\nonumber\\
&= \textrm{Cov}(\tau, \left(f-\overline{f}\right)^2) + \overline{\tau}\sigma^2_{f}\label{cov_to_var_sub_for_variances}
Finally, substituting \eqref{cov_to_var_sub_for_variances} into \eqref{intermediate_3_for_variances}, we see that the rate of change of the variance of any type-level quantity $f$ in the population satisfies:

@@

d\sigma^2_{f} &= \textrm{Cov}\left(w,(f - \overline{f})^2\right)dt\\[12pt]
&- \frac{1}{KN_K}\left[ \ 2\textrm{Cov}\left(\tau,(f - \overline{f})^2\right) + \overline{\tau}\sigma^2_{f}(t) \ \right]dt\\[12pt]

@@

>>>>

&- \frac{1}{KN^2_{K}(t)}\left(\sum\limits_{i=1}^{m}(f_i - \overline{f})^2A_i^+\right)dt\\

@@

\frac{1}{KN^2_{K}(t)}\left(\sum\limits_{i=1}^{m}(f_i-\overline{f})^2A_i^+\right)dt &= \frac{1}{KN_{K}}\sum\limits_{i=1}^{m}(f_i-\overline{f})^2\left(\tau_ip_i + \mu Q_i(\mathbf{p})\right)\\
&= \frac{1}{KN_K}\sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2\left(\tau_ip_i + \mu Q_i(\mathbf{p})\right)\\
&= \frac{1}{KN_K}\left(\overline{\left(f - \overline{f}\right)^2\tau}+\mu \sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2Q_i(\mathbf{p})\right)\\
&= \frac{1}{KN_K}\left(\textrm{Cov}(\tau,\left(f - \overline{f}\right)^2) + \overline{\tau}\overline{\left(f - \overline{f}\right)^2} +\mu \sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2Q_i(\mathbf{p})\right)\\
&= \frac{1}{KN_K}\left(\textrm{Cov}(\tau,\left(f - \overline{f}\right)^2) + \overline{\tau}\sigma^2_{f} +\mu \sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2Q_i(\mathbf{p})\right)

@@

where we have used the definition of statistical covariance in the second to last line and used the definition of statistical variance in the last line. Substituting this into \eqref{intermediate_1_for_variances} and using $M_{\sigma^2_f}(\mathbf{p},N_K)$ to denote the contributions of all the mutational terms (\emph{i.e.} all terms with a $\mu$ factor) for notational brevity, we obtain

@@

d\sigma^2_{f} &= \textrm{Cov}(w,f^2 - 2\overline{f}f)dt - \frac{1}{KN_K}\left(\textrm{Cov}(\tau,f^2 - 2\overline{f}f) + \textrm{Cov}(\tau,\left(f - \overline{f}\right)^2) + \overline{\tau}\sigma^2_{f}\right)dt\\
&+ 2\textrm{Cov}\left(f, \frac{\partial f}{\partial t}\right)dt + M_{\sigma^2_f}(\mathbf{p},N_K)dt + dB_{\sigma^2_{f}}

@@

&- \frac{1}{KN_K}\left[ \ \textrm{Cov}\left(\tau,(f - \overline{f})^2\right) - \textrm{Cov}\left(\tau, {\left(\overline{f}\right)}^2\right) + \textrm{Cov}(\tau,\left(f - \overline{f}\right)^2) + \overline{\tau}\sigma^2_{f} \ \right]dt\\[12pt]

@@

&= {\left(\overline{f}\right)}^2\overline{w} - \overline{w}{\left(\overline{f}\right)}^2 = 0
and similarly,
\begin{equation*}
\textrm{Cov}\left(\tau, {\left(\overline{f}\right)}^2\right) = 0
\end{equation*}
and thus, using this in \eqref{intermediate_2_for_variances},  we see that the rate of change of the variance of any type-level quantity $f$ in the population satisfies:

@@

\label{stochastic_Price_variance}

@@

d\sigma^2_{f} &= \textrm{Cov}\left(w,(f - \overline{f})^2\right)dt - \frac{1}{KN_K}\left[ \ \overline{\tau}\sigma^2_{f} +  2\textrm{Cov}\left(\tau,(f - \overline{f})^2\right) \ \right]dt\\[12pt]

@@

In the case of one-dimensional quantitative traits, an infinite-dimensional version of \eqref{stochastic_Price_variance} has been rigorously derived \citep{week_white_2021} in a recent paper for the case where $f$ is the phenotypic value of the quantitative trait - Equation (21b) of \cite{week_white_2021} is precisely the $m \to \infty$ version of my equation \eqref{stochastic_Price_variance} when $f$ is the phenotypic value of the quantitative trait under consideration.\\
Once again, terms of equation \eqref{stochastic_Price_variance} lend themselves to straightforward biological interpretation. The quantity $(f-\overline{f})^2$ is a measure of the `spread' of the quantity $f$ around the population mean value $\overline{f}$, and thus covariance with $(f-\overline{f})^2$ is a measure of disruptive (\emph{i.e.} diversifying) selection: If values of $f$ that are further away from the mean value are associated with higher fitness, the variance of that quantity in the population increases. The $\overline{\tau}\sigma^2_{f}$ term represents a loss of diversity due to stochastic extinctions (i.e. demographic stochasticity). To see this, let us consider the case in which each $f_i$ is simply a label or mark arbitrarily assigned to individuals in the population at the start of an experiment/observation and subsequently passed on to offspring - For example, a neutral tag in the genome of the organism in an area that does not mutate. Let us set $\mu = 0$ so that the labels cannot change between parents to offspring. This means that we have $M_{\sigma^2_f}(\mathbf{p},N_K) \equiv 0$. Further, since the labels are arbitrary and have no effect whatsoever on the biology of the organisms, we have $\textrm{Cov}\left(w,(f - \overline{f})^2\right) \equiv \textrm{Cov}\left(\tau,(f - \overline{f})^2\right) \equiv 0$. Since the labels do not change over time, we also have $\textrm{Cov}\left(\frac{\partial f}{\partial t},f\right) = 0$. From \eqref{stochastic_Price_variance}, we see that in this case, the variance changes as

@@

d\sigma^2_f = - \frac{\overline{\tau}\sigma^2_{f}}{KN_K}dt + dB_{\sigma^2_{f}}

@@

In particular, on taking expectations, we see that the expected variance in the population obeys
\frac{d \mathbb{E}[\sigma^2_f]}{dt} = - \left(\mathbb{E}\left[\frac{\overline{\tau}}{KN_K}\right]\right)\mathbb{E}[\sigma^2_{f}]
We were able to decompose the expectation on the RHS because the label $f$ is completely arbitrary and thus independent of both $\overline{\tau}$ and $N_K(t)$. This is a simple linear ODE and has the solution
\mathbb{E}[\sigma^2_f(t)] = \sigma^2_f(0)\exp\left(-\mathbb{E}\left[\frac{\overline{\tau}}{KN_K}\right]\right)
In this case, demographic stochasticity is expected to lead to an exponential loss of diversity in the population. Populations with higher mean turnover $\overline{\tau}$  and/or lower population size $KN_K$ lose diversity faster, which again should be intuitive since these populations experience more extreme stochastic events and are therefore more likely to experience loss of labels. Note that \emph{which} label is lost is entirely random (since labelling is arbitrary), but nevertheless, labels tend to be stochastically lost until only a small number of labels remain in the population.
